---
title: "SIREN Cox models - R implementation"
author: 
- "Peter Kirwan, MRC Biostatistics Unit, University of Cambridge"
date:   "`r Sys.Date()`"
output: 
    html_document:
        code_folding: hide
        toc:          true
---

```{r, include=FALSE}

#libraries
library(tidyverse)
library(survival)

# load data
load("~/coviddata/siren_post_processing.RData")

```

```{r labels-functions}

# define labels
ve_label <- tibble(label = c("Unvaccinated",
                       
                       "Pfizer dose 1, 0-20 days", 
                       "Pfizer dose 1, 21-27 days", 
                       "Pfizer dose 1, 28-41 days", 
                       "Pfizer dose 1, 42-55 days", 
                       "Pfizer dose 1, >55 days",
                       
                       "ChAdOx dose 1, 0-20 days", 
                       "ChAdOx dose 1, 21-27 days", 
                       "ChAdOx dose 1, 28-41 days", 
                       "ChAdOx dose 1, 42-55 days", 
                       "ChAdOx dose 1, >55 days",
                       
                       "Pfizer long dose 2, 0-13 days", 
                       "Pfizer long dose 2, 14-73 days", 
                       "Pfizer long dose 2, 74-133 days", 
                       "Pfizer long dose 2, 134-193 days", 
                       "Pfizer long dose 2, >193 days", 
                       
                       "Pfizer short dose 2, 0-13 days", 
                       "Pfizer short dose 2, 14-73 days", 
                       "Pfizer short dose 2, 74-133 days", 
                       "Pfizer short dose 2, 134-193 days", 
                       "Pfizer short dose 2, >193 days", 
                       
                       "ChAdOx dose 2, 0-13 days", 
                       "ChAdOx dose 2, 14-73 days", 
                       "ChAdOx dose 2, 74-133 days", 
                       "ChAdOx dose 2, >133 days"
                       ),
             group = rep("Vaccine, dose, schedule",25))

dura_label <- tibble(label = c("Unvaccinated, naive", 
                        "Unvaccinated, <3 months", 
                        "Unvaccinated, 3-9 months", 
                        "Unvaccinated, 9-15 months", 
                        "Unvaccinated, >15 months", 
                        "Dose 1, naive", 
                        "Dose 1, <3 months", 
                        "Dose 1, 3-9 months", 
                        "Dose 1, 9-15 months", 
                        "Dose 1, >15 months", 
                        "Dose 2, naive", 
                        "Dose 2, <3 months", 
                        "Dose 2, 3-9 months", 
                        "Dose 2, 9-15 months", 
                        "Dose 2, >15 months"
                        ),
                 group = rep("Vaccine, prior infection",15))

labels_all <- tibble(label = c(
                 "East Midlands",
                 "East of England", 
                 "London", 
                 "North East", 
                 "North West", 
                 "South East", 
                 "South West", 
                 "West Midlands", 
                 "Yorkshire and Humber",
                 "Scotland",
                 "Northern Ireland",
                 "Wales",
                 "<25",
                 "25-34",
                 "35-44",
                 "45-54",
                 "55-64",
                 "65+",
                 "Unknown age",
                 "Male",
                 "Female",
                 "Non-binary",
                 "Unknown gender",
                 "White",
                 "Mixed",
                 "Asian",
                 "Black",
                 "Other ethnicity",
                 "Unknown ethnicity",
                 "Every day",
                 "Once a week",
                 "Once a month",
                 "Less than once a month",
                 "Never",
                 "Office",
                 "Patient facing (non-clinical)",
                 "Outpatient",
                 "Maternity,Labour Ward",
                 "Ambulance,Emergency Department,Inpatient",
                 "Intensive care",
                 "Theatres",
                 "Other"),
                 group = c(rep("Region",12), 
                           rep("Age group",7), 
                           rep("Gender",4), 
                           rep("Ethnicity",6), 
                           rep("Work exposure",5),
                           rep("Occupation", 8))
                 )

# define functions
results_table <- function(cox_model, title){
    
    labels %>%
        full_join(
            tidy_siren_cox %>%
                bind_cols(levels %>% select(label)),
            by="label") %>%
        mutate(estimate = if_else(is.na(estimate),1,estimate)) %>% 
        select(label, estimate, conf.low, conf.high) %>% 
        knitr::kable(caption = title)
}

results_figure <- function(cox_model, title){
    
    labels %>%
        full_join(
            tidy_siren_cox %>%
                bind_cols(levels %>% select(label)),
            by="label") %>%
        mutate(estimate = if_else(is.na(estimate),1,estimate),
               label = factor(label, levels = labels$label),
               ref=if_else(estimate==1,"ref","not ref")) %>%
        ggplot()+
        aes(y=label, x=estimate, xmin = conf.low, xmax = conf.high, shape = ref, color = group)+
        geom_point(size=1.5)+
        geom_errorbarh(height=.3)+
        scale_x_continuous(name="Hazard ratio", trans='log10')+
        scale_y_discrete(limits=rev, name="")+
        geom_vline(xintercept=1, color="black", linetype="dashed", alpha=.5)+
        labs(subtitle = title,
             caption = "Shown on log scale, hollow points indicate reference group") +
        theme_minimal()+
        scale_shape_manual(values = c(19, 1))+
        theme(legend.position = "none")+
        facet_grid(group ~ ., space="free_y", scales="free_y", switch="y")+
        theme(panel.spacing = unit(1, "lines"))+
        theme(strip.placement = "outside",
              strip.background = element_rect(fill=NA,colour=NA),
              panel.spacing=unit(0.3,"cm"), axis.title.y = element_blank())+
        theme(text = element_text(size = 13))
}

```

This document presents Cox Proportional Hazards estimates for the SIREN interim analysis dataset (SIREN_Interim_20210921_v2.dta). This analysis replicates a set of Stata analyses, thereby providing independent verification.

Survival analysis datasets with multiple intervals for each participant and corresponding covariate values were prepared using the `survival` package (Therneau T (2021). _A Package for Survival Analysis in R_.)

Minor differences exist between the Stata and `survival` package implementations of the Cox Proportional Hazards model:

- The default method for handling tied event times in Stata is the Breslow approximation, whereas in the `survival` package the more accurate Efron approximation is the default. We decided to use the Efron approximation.

- Clustering on NHS trust is used for robust variance calculation. When using clustering, Stata makes an adjustment to the standard errors - multiplying by g/(g − 1), where g is the number of clusters - whereas the `survival` package does not make this adjustment. We decided to use the adjusted standard errors.

## 1. Vaccine effectiveness

The vaccination categories were: unvaccinated, 0/21/28/42/56 days post-first dose and 0/14/74/134/194 days post-second dose.

These categories were further sub-divided according to vaccine: Pfizer long, Pfizer short, and ChAdOx.

Follow-up time for those within 90 days of a primary infection was removed, as was the category "ChAdOx dose 2, 0-13 days" due to insufficient observations.

```{r ve, warning=FALSE}

# vaccine effectiveness (unadjusted)
siren_cox <- coxph(Surv(tstart, tstop, event) ~ vaccine_cat, 
                   id=study_id, 
                   cluster=trust_code, 
                   data=split_ve %>% filter(eligible==1, vaccine_cat!=21)
                   )

tidy_siren_cox <- broom::tidy(siren_cox, conf.int = TRUE, exponentiate = TRUE)

labels <- ve_label
levels <- labels %>% filter(!label %in% c("Unvaccinated"))

results_table(tidy_siren_cox, title = "Unadjusted model")
results_figure(tidy_siren_cox, title = "Unadjusted model")

```

```{r ve-adjusted, fig.width=10, fig.height=10, warning=FALSE}

# vaccine effectiveness (adjusted)
siren_cox <- coxph(Surv(tstart, tstop, event) ~ vaccine_cat + primary_inf + region + agegr2 + gender + ethnic_gr + work_exposure_frequency + occ_set_cat,
                   id=study_id, 
                   cluster=trust_code, 
                   data=split_ve %>% filter(eligible==1, vaccine_cat!=21)
                   )

tidy_siren_cox <- broom::tidy(siren_cox, conf.int = TRUE, exponentiate = TRUE)

labels <- ve_label %>% 
    rbind(tibble(label=c("Naive", "0-3 months", "3-9 months", "9-15 months",">15 months"), 
                 group=rep("Primary infection",5))) %>%  
    rbind(labels_all)

levels <- labels %>% filter(!label %in% c("Unvaccinated","Naive", "East Midlands","<25","Male","White","Every day","Office"))

results_table(tidy_siren_cox, title = "Adjusted model")
results_figure(tidy_siren_cox, title = "Adjusted model")

# if we wish to match the standard error calculation from Stata:
# sqrt(diag(vcov(siren_cox))* (135/134))

```


## 2. Durability of protection

The vaccination categories were: unvaccinated, first dose and second dose.

The primary infection categories were: naive, 3-9 months, 9-15 months, and >15 months since primary infection.

Those within 90 days of a primary infection were excluded.

```{r durability, warning=FALSE}

# durability of protection (unadjusted)
siren_cox <- coxph(Surv(tstart, tstop, event) ~ primary_cat,
                   id=study_id, 
                   cluster=trust_code, 
                   data=split_dura %>% filter(eligible==1)
                   )

tidy_siren_cox <- broom::tidy(siren_cox, conf.int = TRUE, exponentiate = TRUE)

labels <- dura_label
levels <- labels %>% filter(!label %in% c("Unvaccinated, naive"))

results_table(tidy_siren_cox, title = "Unadjusted model")
results_figure(tidy_siren_cox, title = "Unadjusted model")

```

```{r durability-adjusted, fig.width=10, fig.height=10, warning=FALSE}

# durability of protection (adjusted)
siren_cox <- coxph(Surv(tstart, tstop, event) ~ primary_cat + region + agegr2 + gender + ethnic_gr + work_exposure_frequency + occ_set_cat,
                   id=study_id, 
                   cluster=trust_code, 
                   data=split_dura %>% filter(eligible==1)
                   )

tidy_siren_cox <- broom::tidy(siren_cox, conf.int = TRUE, exponentiate = TRUE)

labels <- dura_label %>% rbind(labels_all)
levels <- labels %>% filter(!label %in% c("Unvaccinated, naive", "East Midlands","<25","Male","White","Every day","Office"))

results_table(tidy_siren_cox, title = "Adjusted model")
results_figure(tidy_siren_cox, title = "Adjusted model")

```

## Session info

```{r session-info, results='asis'}

sessionInfo() %>% 
    pander::pander()

```
