---
title: "SIREN Cox models - R implementation"
author: 
- "Peter Kirwan, MRC Biostatistics Unit, University of Cambridge"
date:   "`r Sys.Date()`"
output: 
    html_document:
        code_folding: hide
        toc:          true
---

```{r, include=FALSE}

# libraries
library(tidyverse)
library(survival)
library(finalfit)
library(flextable)

# load data
load("~/coviddata/siren_post_processing.RData")

```

```{r labels-functions}

# define labels
ve_label <- c(
    "Unvaccinated",
    "Pfizer dose 1, 0-20 days",
    "Pfizer dose 1, 21-27 days",
    "Pfizer dose 1, 28-41 days",
    "Pfizer dose 1, 42-55 days",
    "Pfizer dose 1, >55 days",
    "ChAdOx dose 1, 0-20 days",
    "ChAdOx dose 1, 21-27 days",
    "ChAdOx dose 1, 28-41 days",
    "ChAdOx dose 1, 42-55 days",
    "ChAdOx dose 1, >55 days",
    "Pfizer long dose 2, 0-13 days",
    "Pfizer long dose 2, 14-73 days",
    "Pfizer long dose 2, 74-133 days",
    "Pfizer long dose 2, 134-193 days",
    "Pfizer long dose 2, >193 days",
    "Pfizer short dose 2, 0-13 days",
    "Pfizer short dose 2, 14-73 days",
    "Pfizer short dose 2, 74-133 days",
    "Pfizer short dose 2, 134-193 days",
    "Pfizer short dose 2, >193 days",
    "ChAdOx dose 2, 0-13 days",
    "ChAdOx dose 2, 14-73 days",
    "ChAdOx dose 2, 74-133 days",
    "ChAdOx dose 2, >133 days"
)

dura_label <- c(
    "Unvaccinated, naive",
    "Dose 1 0-20 days, naive",
    "Dose 1 21+ days, naive",
    "Dose 2 0-13 days, naive",
    "Dose 2 14-73 days, naive",
    "Dose 2 73-133 days, naive",
    "Dose 2 133-193 days, naive",
    "Dose 2 193+ days, naive",
    "Unvaccinated, <1yr",
    "Dose 1 0-20 days, <1yr",
    "Dose 1 21+ days, <1yr",
    "Dose 2 0-13 days, <1yr",
    "Dose 2 14-73 days, <1yr",
    "Dose 2 73-133 days, <1yr",
    "Dose 2 133-193 days, <1yr",
    "Dose 2 193+ days, <1yr",
    "Unvaccinated, >1yr",
    "Dose 1 0-20 days, >1yr",
    "Dose 1 21+ days, >1yr",
    "Dose 2 0-13 days, >1yr",
    "Dose 2 14-73 days, >1yr",
    "Dose 2 73-133 days, >1yr",
    "Dose 2 133-193 days, >1yr",
    "Dose 2 193+ days, >1yr"
)

```

This document presents Cox Proportional Hazards estimates for the SIREN interim analysis dataset (SIREN_Interim_20210921_v2.dta). This analysis replicates a set of Stata analyses, thereby providing independent verification.

Survival analysis datasets with multiple intervals for each participant and corresponding covariate values were prepared using the `survival` package (_Therneau T_) and formatted using the `finalfit` package (_Harrison E, Drake T, and Ots R_).

Minor differences exist between the Stata and `survival` package implementations of the Cox Proportional Hazards model:

- The default method for handling tied event times in Stata is the Breslow approximation, whereas in the `survival` package the more accurate Efron approximation is the default. We decided to use the Efron approximation.

- Clustering on NHS trust is used for robust variance calculation. When using clustering, Stata makes an adjustment to the standard errors - multiplying by g/(g − 1), where g is the number of clusters - whereas the `survival` package does not make this adjustment. We decided to use the adjusted standard errors.

## 1. Vaccine effectiveness

The vaccination categories were: unvaccinated, 0/21/28/42/56 days post-first dose, and 0/14/74/134/194 days post-second dose.

These categories were further sub-divided according to vaccine: Pfizer long, Pfizer short, and ChAdOx.

Follow-up time for those within 90 days of a primary infection was removed, as was the category "ChAdOx dose 2, 0-13 days" due to insufficient observations.

```{r ve, fig.width=10, warning=FALSE, message=FALSE}

# define variables
dependent = "Surv(tstart, tstop, event)"
explanatory_uni = "vaccine_cat"
explanatory_multi = c("vaccine_cat", "gender", "ethnic_gr", "strata(region)", "strata(agegr2)", "strata(work_exposure_frequency)", "strata(occ_set_cat)", "cluster(trust_code)")

# format data
model_data <- split_ve %>% mutate(vaccine_cat=factor(vaccine_cat, labels = ve_label)) %>% 
    filter(eligible == 1, primary_inf==0, vaccine_cat !="ChAdOx dose 2, 0-13 days") %>% 
    mutate(vaccine_cat = fct_drop(vaccine_cat))

# kaplan meier survival curves
survfit(as.formula(paste0(dependent, "~", explanatory_uni)), 
                  data = model_data) %>% 
    broom::tidy() %>% 
    ggplot()+
    aes(time, estimate, color = strata)+
    geom_step()+
    theme_minimal()

# test proportional hazards assumption
# not significant - indicates proportional hazards assumption holds
model_data %>% 
    coxphmulti(dependent, explanatory_uni) %>% 
    cox.zph()

model_data %>% 
    coxphmulti(dependent, explanatory_multi) %>% 
    cox.zph()

# table, finalfit provides both unadjusted and adjusted hazards
model_data %>% finalfit(dependent, explanatory_multi) %>% 
    flextable() %>% 
    autofit()

# if we wish to match the standard error calculation from Stata:
# sqrt(diag(vcov(siren_cox))* (135/134))

```

```{r ve-fig, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}

# figure
model_data %>% 
    hr_plot(dependent, explanatory_multi, dependent_label = "Survival")

```


## 2. Durability of protection

The vaccination categories were: unvaccinated, 0/21 days post-first dose, and 0/14/74/134/194 days post-second dose.

The primary infection categories were: naive, <1 year, and >1 year since primary infection.

Follow-up time for those within 90 days of a primary infection was removed, as was the category "Dose 1 0-20 days, >1yr" due to insufficient observations.

```{r dura, fig.width=10, warning=FALSE, message=FALSE}

# durability of protection (unadjusted)
dependent = "Surv(tstart, tstop, event)"
explanatory_uni = "primary_cat"
explanatory_multi = c("primary_cat", "strata(region)", "strata(agegr2)", "gender", "ethnic_gr", "strata(work_exposure_frequency)", "strata(occ_set_cat)", "cluster(trust_code)")

# format data
model_data <- split_dura %>% 
    mutate(primary_cat=factor(primary_cat, labels = dura_label)) %>% 
    filter(eligible == 1,  (is.na(vaccine_name2) | vaccine_name2!=2), primary_cat != "Dose 1 0-20 days, >1yr") %>% 
    mutate(primary_cat = fct_drop(primary_cat))

# kaplan meier survival curves
survfit(as.formula(paste0(dependent, "~", explanatory_uni)), 
                  data = model_data) %>% 
    broom::tidy() %>% 
    ggplot()+
    aes(time, estimate, color = strata)+
    geom_step()+
    theme_minimal()

# test proportional hazards assumption
# significant - indicates proportional hazards assumption does not hold
model_data %>% 
    coxphmulti(dependent, explanatory_uni) %>% 
    cox.zph()

model_data %>% 
    coxphmulti(dependent, explanatory_multi) %>% 
    cox.zph()

# table, finalfit provides both unadjusted and adjusted hazards
model_data %>% finalfit(dependent, explanatory_multi) %>% 
    flextable() %>% 
    autofit()

```

```{r dura-fig, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}

# figure
model_data %>% 
    hr_plot(dependent, explanatory_multi, dependent_label = "Survival")

```

## Session info

```{r session-info, results='asis'}

sessionInfo() %>%
    pander::pander()
```
